Smoothing seismicity techniques applied to seismic sources characterization and probabilistic seismic hazard analysis in Brazil

### 2015-01-14 manuscript.tex
methodology

\documentclass[draft, grl]{agutex}

@@

% \usepackage{lineno}
% \linenumbers*[1]

@@

% Glossaries
\usepackage[sanitize=none,acronym,toc]{glossaries}
%\usepackage[acronym,toc]{glossaries}
% Define a new glossary type
\newglossary[slg,toc]{symbols}{sym}{sbl}{Lista de Símbolos}
\newglossary[elg]{equations}{eqn}{eql}{Equations}
\makeglossaries

@@

Frankel in their method uses a \emph{fixed} bandwidth in space. Woo's method
suggest a bandwidth as function of the \emph{magnitude}.
Helmstetter and Werner proposes a \emph{local-adaptive}
in space and time kernel bandwidths, chosen for each event.
This last one was proposed initialy to take care of the background seismicity rate
on long-term earthquake forecast context and the time-independent assumption
about the annual background seismic rate could be used
(not indiscriminately REFERENCE) as a grid of points seismic sources.

@@

%----------------------------------------------------------------------------------------
%	ARTICLE CONTENT
%----------------------------------------------------------------------------------------

% The body of the article must start with a \begin{article} command
% \end{article} must follow the references section, before the figures and tables.

@@

Although the standard Cornell Mcguire zoning methodolgy for the classical hazard seismic assesment, the goal is
review a few smoothing seismicity methods based on kernel density estimation. This estimation allow a seismic point
sources grid definition. The seismic rate (a-value) associated to each source is definided by the smoothed seismicity
methods, but the more parameters like b-values, minimum and maximum magnitudes, rupture and depth
distributions must to be assigned in other ways.

@@

%------------------------------------------------
\section{Methods}

@@

Gordon \citet{woo_1996} suggested to evaluate the contribution of each earthquake $i$, far $\boldsymbol{r}_i$ from its kernel, on the cell centered on $\boldsymbol{r}$ in dependence of their magnitude $m$:
\begin{equation}
	\ensuremath{
		\gls{sym:Rrm} = \sum_{i=1}^{N} \frac{ K(\gls{sym:r} - \gls{sym:ri}, m)}
											{T({\gls{sym:ri})}},
	}
	\label{eq:Rrm}
\end{equation}
where $N$ is the number of earthquakes $i$ on the catalogue
and $T(\gls{sym:ri})$ is the timeframe which all earthquakes with magnitude above $m$ is completely observed on \gls{sym:ri}.

Any kernel could be applied on that definition. In practice Woo used a \citet{kagan_knopoff_1980} for a infinity spatial domain:
\begin{equation}
	\ensuremath{
		K_{KJ}(\gls{sym:r}, m \arrowvert \gls{sym:aW}) =  \frac{  \gls{sym:aW}  -1}{\pi\gls{sym:hm}^2}
							\left[ 1 + \frac{\gls{sym:r}^2}{\gls{sym:hm}^2} \right]^{-\gls{sym:aW}},
	}
	\label{eq:k_kj}
\end{equation}
where \gls{sym:aW} is, accordingly \citet{verejones_1992}, \glsdesc{sym:aW}.

@@

\subsection{Helmstetter}

\begin{equation}
	\ensuremath{\gls{sym:R} = \sum_{i=1}^{N}{ \frac{1}{h_i\,{d_i}^2} \gls{sym:Kt}\gls{sym:Kr} }},
	\label{eq:helms01}
\end{equation}
onde \gls{sym:R} é \glsdesc{sym:R},
	  $K_t$ é a \glsdesc{sym:Kt},
	  $K_r$ é a \glsdesc{sym:Kr}.

Como seu interesse é em projeção, restringiram os tempos do catálogo aos $t_i
< t$ e ponderaram segundo um peso $w$ da seguinte maneira:
\begin{equation}
\ensuremath{\gls{sym:R} = \gls{sym:Rmin} + \sum_{t_i < t}{
	\frac{2\,w(\boldsymbol{r}_i,t_i)}{h_i\,{d_i}^2}
			\gls{sym:Kt}\gls{sym:Kr} }},
	\label{eq:helms02}
\end{equation}
onde \gls{sym:Rmin} é a \glsdesc{sym:Rmin}, positiva, permitindo que ocorram eventos-surpresa
onde não se tem registro de sismicidade.

Os pesos $w(\boldsymbol{r}_i,t_i)$, calculados em cada tremor $i$, propostos por Helmstetter são a transformação de
Gutemberg-Richter. Consideraram uma contribuição maior para a taxa de sismicidade daqueles sismos
que ocorreram onde e quando a magnitude de completude $M_c$ era maior que
a mínima magnitude $M_d$ do catálogo
em locais onde a magnitude de completude e o \emph{valor-b} variam.
Note que o modelo de pesos acomoda bem variações espaciais e
temporais tanto da magnitude de completude como do \emph{valor-b}. A expressão é a seguinte:
\begin{equation}
	\ensuremath{ w(\boldsymbol{r},t) = 10^{ \gls{sym:b}(\boldsymbol{r},t) \left[ \gls{sym:Mc_rt} - \gls{sym:Md}
	\right] } },
	\label{eq:helms_wi}
\end{equation}
onde  \gls{sym:wi} é o \glsdesc{sym:wi} na localização $\boldsymbol{r}$ e no instante $t$,
	  \gls{sym:b}$(\boldsymbol{r},t)$ é o \glsdesc{sym:b},
	  \gls{sym:Mc_rt} é a \glsdesc{sym:Mc_rt},
	  \gls{sym:Md} é a \glsdesc{sym:Md}.

@@

\subsubsection{Taxa de sismicidade estacionária}
Mesmo havendo modelado a dependência com o tempo para distribuição da taxa de sismicidade,
a taxa estacionária $\bar{R}$ que se preserva no longo-prazo em uma determinada localização
$\boldsymbol{r}_0$ é a que mais interessa à \gls{psha} e,
segundo o método proposto por \citet{helmstetter_2012}, será definida
pelo valor da mediana da taxa fornecida pelo modelo no
período considerado:

\begin{equation}
	\ensuremath{
		\bar{R}(\boldsymbol{r}_0) = \text{Mediana}\left[R(\boldsymbol{r}_0, t)\right].
	}
	\label{eq:stationary_rate}
\end{equation}

O fato de se tomar a mediana faz com que variações na taxa de sismicidade devido a pré e pós-eventos
não interfiram significativamente no valor da taxa estacionária. A decorrência mais importante desse fato
é que o método dispensa a remoção de agrupamentos (seção \ref{sec:declustering}) do catálogo, um método paramétrico e
quase sempre controverso.

\subsubsection{Verossimilhança}

Para otimizar os parâmetros do modelo de Helmstetter, é preciso dividir o catálogo de sismos.
Uma parte serve para o aprendizado do modelo e a outra para avaliação.

Se a ocorrência de tremores pode ser modelada por um processo de Poisson com taxa $N_p$,
então a probabilidade de se observar exatamente $n$ eventos no período de tempo considerado é dada por

\begin{equation}
	\ensuremath{
		\gls{sym:pNn} = \frac{{N_p}^n e^{-N_p}}{n!}.
	}
	\label{eq:poisson_probability}
\end{equation}

E o logarítmo, a ser maximizado, da verossimilhança entre
o que o modelo predisse e o que foi observado
pode ser escrito como

\begin{equation}
	\ensuremath{
		\gls{sym:L} = \sum_{i_x=1}^{N_x}\sum_{i_y=1}^{N_y}\log p\left[  \gls{sym:Np}, \gls{sym:nxy}  \right]
	}
	\label{eq:loglik}
\end{equation}
onde \gls{sym:Np} é \glsdesc{sym:Np},
	\gls{sym:nxy} é \glsdesc{sym:nxy}.

Os parâmetros $R_{min}$, $a_{cnn}$ e $k_{cnn}$ do modelo são otimizados
pela maximização da verossimilhança $L$.

@@

Referencing equation \ref{eq:emc}. Mauris vel lorem magna, tristique auctor ipsum. Aliquam pharetra eleifend massa. Donec porttitor sagittis luctus. Aliquam pretium luctus leo quis congue. Morbi vel felis mi. Referencing Table \ref{fig_seismicity}. Referencing Figure \ref{placeholder}.

@@

%----------------------------------------------------------------------------------------
%	ACKNOWLEDGEMENTS
%----------------------------------------------------------------------------------------

\begin{acknowledgments}
The author thanks specially to Marcelo Assumpção for the support and critical discussion as well all IAG-USP seismological team. It is also grateful to Stéphane Drouet and Vincent Guigues for all helpful discussions.
\end{acknowledgments}

%----------------------------------------------------------------------------------------
%	BIBLIOGRAPHY
%----------------------------------------------------------------------------------------

@@

% ---------------------------------------------------------------------------- %
% Bibliografia
\bibliographystyle{BibTeX/agufull08} 	% citação bibliográfica textual
\bibliography{bib/bibliografia}  		% associado ao arquivo:

@@

%----------------------------------------------------------------------------------------
%	FIGURES AND TABLES
%----------------------------------------------------------------------------------------

@@

\begin{figure}
	\includegraphics[width=0.4\linewidth]{z_img_helmstetter_catalogues}
	\caption{Learning and target catalogues.}
	\label{fig_helmstetter_catalogues}
\end{figure}

@@

% AGU follows standard outline style; therefore, there cannot be a section 1 without
% a section 2, or a section 2.3.1 without a section 2.3.2.
% Please make sure your section numbers are balanced.
% ---------------
% Level 1 head
%
% Use the \section{} command to identify level 1 heads;
% type the appropriate head wording between the curly
% brackets, as shown below.
%
%An example:
%\section{Level 1 Head: Introduction}
%
% ---------------
% Level 2 head
%
% Use the \subsection{} command to identify level 2 heads.
%An example:
%\subsection{Level 2 Head}
%
% ---------------
% Level 3 head
%
% Use the \subsubsection{} command to identify level 3 heads
%An example:
%\subsubsection{Level 3 Head}
%
%---------------
% Level 4 head
%
% Use the \subsubsubsection{} command to identify level 3 heads
% An example:
%\subsubsubsection{Level 4 Head} An example.
%
%% ------------------------------------------------------------------------ %%
%
%  IN-TEXT LISTS
%
%% ------------------------------------------------------------------------ %%
%
% Do not use bulleted lists; enumerated lists are okay.
% \begin{enumerate}
% \item
% \item
% \item
% \end{enumerate}
%
%% ------------------------------------------------------------------------ %%
%
%  EQUATIONS
%
%% ------------------------------------------------------------------------ %%

>>>>

\documentclass[draft, grl]{agutex}
%\documentclass[grl]{agutex}

@@

\usepackage{lineno}
\linenumbers*[1]

@@

% Glossaries
%\usepackage[sanitize=none,acronym,toc]{glossaries}
\usepackage[acronym,toc]{glossaries}
%\usepackage[acronym,toc]{glossaries}
% Define a new glossary type
\newglossary[slg,toc]{symbols}{sym}{sbl}{Lista de Símbolos}
\newglossary[elg]{equations}{eqn}{eql}{Equations}
\makeglossaries

@@

Frankel in their method uses a \emph{fixed} bandwidth in space. Woo's method suggest a bandwidth as function of the \emph{magnitude}. Helmstetter and Werner proposes a \emph{local-adaptive} in space and time kernel bandwidths, chosen for each event.
This last one was proposed initialy to take care of the background seismicity rateon long-term earthquake forecast context and the time-independent assumption about the annual background seismic rate could be used
(not indiscriminately REFERENCE) as a grid of points seismic sources.

@@

%------------------------------------------------------------------------------
%	ARTICLE CONTENT
%------------------------------------------------------------------------------
% The body of the article must start with a \begin{article} command
% \end{article} must follow the references section, before the figures and tables.

@@

Although the standard Cornell Mcguire zoning methodolgy for the classical hazard seismic assesment, the goal is review a few smoothing seismicity methods based on kernel density estimation. This estimation allow a seismic point sources grid definition. The seismic rate (\emph{a}-value) associated to each source is definided by the smoothed seismicity methods, but the more parameters like \emph{b}-values, minimum and maximum magnitudes, rupture and depth distributions must to be assigned in other ways.

@@

%------------------------------------------------
\section{Smoothing methods}
Under this work, three smoothing methods was considered. On this section brief explanation of each one will be done.

@@

Gordon \citet{woo_1996} suggested to evaluate the contribution of each earthquake $i$, far $\boldsymbol{r}_i$ from its kernel, on the cell centered on $\boldsymbol{r}$ in dependence of their magnitude $m$:
\begin{equation}
	\ensuremath{
		\gls{sym:Rrm} = \sum_{i=1}^{N} \frac{ K(\gls{sym:r} - \gls{sym:ri}, m)}
											{T({\gls{sym:ri})}},
	}
	\label{eq:Rrm}
\end{equation}
where $N$ is the number of earthquakes $i$ on the catalog
and $T(\gls{sym:ri})$ is the timeframe which all earthquakes with magnitude above $m$ is completely observed on \gls{sym:ri}.

Any kernel could be applied on that definition. In practice Woo used a \citet{kagan_knopoff_1980} for a infinity spatial domain:
\begin{equation}
	\ensuremath{
		K_{KK}(\gls{sym:r}, m \arrowvert \gls{sym:aW}) =  \frac{  \gls{sym:aW}  -1}{\pi\gls{sym:hm}^2}
							\left[ 1 + \frac{\gls{sym:r}^2}{\gls{sym:hm}^2} \right]^{-\gls{sym:aW}},
	}
	\label{eq:k_kj}
\end{equation}
where \gls{sym:aW} is, accordingly \citet{verejones_1992}, \glsdesc{sym:aW}.

@@

\subsection{Helmstetter and Werner}

Even work on a forecast perspective, the background seismicity rate for a long-term time-independent forecast could be used to charachterize a diffused seismicity under the common assumption that this background seismic rate is invariant over time.

\citet{helmstetter_2012} proposes a seismicity model space and time dependent using kernels for space and time independently:
\begin{equation}
	\ensuremath{\gls{sym:R} = \sum_{i=1}^{N}{ \frac{1}{h_i\,{d_i}^2} \gls{sym:Kt}\gls{sym:Kr} }},
	\label{eq:helms01}
\end{equation}
where \gls{sym:R} is the \glsdesc{sym:R},
	  $K_t$ is the \glsdesc{sym:Kt},
	  $K_r$ is the \glsdesc{sym:Kr}.

As they are interested in forecast, just past time $t_i < t$ need to be considered. Also the observation completeness is taken in account by a set of weights $w$ in this follow way:
\begin{equation}
\ensuremath{\gls{sym:R} = \gls{sym:Rmin} + \sum_{t_i < t}{
	\frac{2\,w(\boldsymbol{r}_i,t_i)}{h_i\,{d_i}^2}
			\gls{sym:Kt}\gls{sym:Kr} }},
	\label{eq:helms02}
\end{equation}
where \gls{sym:Rmin} is the \glsdesc{sym:Rmin}, positive, allowing earthquakes to occur where their never occurred yet.

The weights $w(\boldsymbol{r}_i,t_i)$, computed for each earthquake $i$, are the Gutemberg-Richter \emph{a}-value projection. These weights increase the seismicity contribution from earthquakes occurred where and when the completeness magnitude $M_c$ was greather than minimum catalog magnitude $M_d$. These weights could easily consider space and time completenes and \emph{b}-value fluctuations. The expression used to compute it:
\begin{equation}
	\ensuremath{ w(\boldsymbol{r},t) = 10^{ \gls{sym:b}(\boldsymbol{r},t) \left[ \gls{sym:Mc_rt} - \gls{sym:Md}
	\right] } },
	\label{eq:helms_wi}
\end{equation}
where \gls{sym:wi} is the \glsdesc{sym:wi} at location $\boldsymbol{r}$ on instant $t$,
	  \gls{sym:b}$(\boldsymbol{r},t)$ is the space and time dependent \glsdesc{sym:b},
	  \gls{sym:Mc_rt} is the \glsdesc{sym:Mc_rt},
	  \gls{sym:Md} is the \glsdesc{sym:Md}.

@@

\subsubsection{Stationary seismic rate}
After compute the model parameters and completely define it, the time-independent or stationary seismic rate $\bar{R}$ on each location $\boldsymbol{r}_0$ have to be computed. 
The way to do it proposed by \citet{helmstetter_2012} was to get the median
on $\boldsymbol{r}_0$ over all considered time window:
\begin{equation}
	\ensuremath{
		\bar{R}(\boldsymbol{r}_0) = \text{Median}\left[R(\boldsymbol{r}_0, t)\right].
	}
	\label{eq:stationary_rate}
\end{equation}

The median should avoid the seismicity rate fluctuations derived by fore and after-shocks and consequently the decluster procedures.

\subsubsection{Maximum likelyhood optimization}

The model is completely defined by \gls{sym:k_cnn}, \gls{sym:a_cnn} and \gls{sym:Rmin}. To optimize these parameters the catalog is divided in two parts: one for learning about the parameters and other to test them.

If the earthquake occurrence could be modelled by a Poisson process with rate $N_p$, then the probability to observe exactly $n$ events on the consider time frame is
\begin{equation}
	\ensuremath{
		\gls{sym:pNn} = \frac{{N_p}^n e^{-N_p}}{n!}.
	}
	\label{eq:poisson_probability}
\end{equation}

Then, over all cells, the (log) likelihood, to be maximized, between model (from the learning catalog) prediction and earthquakes observed on testing catalog is written as
\begin{equation}
	\ensuremath{
		\gls{sym:L} = \sum_{i_x=1}^{N_x}\sum_{i_y=1}^{N_y}\log p\left[  \gls{sym:Np}, \gls{sym:nxy}  \right]
	}
	\label{eq:loglik}
\end{equation}
where \gls{sym:Np} is the \glsdesc{sym:Np} and
	  \gls{sym:nxy} is the \glsdesc{sym:nxy}.

The model parameters $R_{min}$, $a_{cnn}$ and $k_{cnn}$ should be optimized by the maximization of log-likelihood $L$.

@@

Referencing equation. Mauris vel lorem magna, tristique auctor ipsum. Aliquam pharetra eleifend massa. Donec porttitor sagittis luctus. Aliquam pretium luctus leo quis congue. Morbi vel felis mi. Referencing Table \ref{fig_seismicity}. Referencing Figure \ref{placeholder}.

@@

%------------------------------------------------------------------------------
%	ACKNOWLEDGEMENTS
%------------------------------------------------------------------------------
\begin{acknowledgments}
The author thanks specially to Marcelo Assumpção for the support and critical discussion as well all IAG-USP seismological team. It is also grateful to Stéphane Drouet and Vincent Guigues for all helpful discussions.
\end{acknowledgments}

%------------------------------------------------------------------------------
%	BIBLIOGRAPHY
%------------------------------------------------------------------------------

@@

% ---------------------------------------------------------------------------- % Bibliografy
% ---------------------------------------------------------------------------- 
\bibliographystyle{BibTeX/agufull08} 	% citação bibliográfica textual
\bibliography{bib/bibliografia}  		% associado ao arquivo:

@@

%------------------------------------------------------------------------------
%	FIGURES AND TABLES
%------------------------------------------------------------------------------

@@

\begin{figure}
	\includegraphics[width=0.4\linewidth]{z_img_helmstetter_catalogues}
	\caption{Learning and target catalogs.}
	\label{fig_helmstetter_catalogues}
\end{figure}

@@

% AGU follows standard outline style; therefore, there cannot be a section 1 without
% a section 2, or a section 2.3.1 without a section 2.3.2.
% Please make sure your section numbers are balanced.
% ---------------
% Level 1 head
%
% Use the \section{} command to identify level 1 heads;
% type the appropriate head wording between the curly
% brackets, as shown below.
%
%An example:
%\section{Level 1 Head: Introduction}
%
% ---------------
% Level 2 head
%
% Use the \subsection{} command to identify level 2 heads.
%An example:
%\subsection{Level 2 Head}
%
% ---------------
% Level 3 head
%
% Use the \subsubsection{} command to identify level 3 heads
%An example:
%\subsubsection{Level 3 Head}
%
%---------------
% Level 4 head
%
% Use the \subsubsubsection{} command to identify level 4 heads
% An example:
%\subsubsubsection{Level 4 Head} An example.
%
%% ------------------------------------------------------------------------ %%
%
%  IN-TEXT LISTS
%
%% ------------------------------------------------------------------------ %%
%
% Do not use bulleted lists; enumerated lists are okay.
% \begin{enumerate}
% \item
% \item
% \item
% \end{enumerate}
%
%% ------------------------------------------------------------------------ %%
%
%  EQUATIONS
%
%% ------------------------------------------------------------------------ %%
